---
title: "App User Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{App User Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{css style settings, echo = FALSE}
blockquote {
    padding: 10px 20px;
    margin: 0 0 20px;
    font-size: 14px;
    border-left: 5px solid;
}
```

# App Overview

While the package contains a number of functions that are useful for correcting sonde data, the main part of this package is the Shiny app which allows the user to interactively load, view, correct, and export sonde data. The app works by creating and modifying a sonde project (`sondeproj`) object which is a list containing sonde data, flags, and metadata. See `help(example_sondeproj)` for more details on project structure.

**Note: This app will not modify any raw data files. It is only able to read raw files and save to specified location.**

The app consists of nine modules or pages. *While they are put in an order that represent a typical correction workflow, the steps do not need to be performed in order.*

1.  **Load Data:** Load either an existing sonde project, load raw data and metadata to create a sonde project, or amend an existing sonde project with new data. Can also load or download precipitation data to include with the project.

2.  **Visualize:** Explore the loaded data via an interactive plot, view metadata or data summary statistics. Can also view data versions and remove periods where the sonde was out of the water.

3.  **Data Checks:** View summaries of data gaps and duplicates. Resolve duplicates and add notes about duplicate/missing data.

4.  **Physical Limits:** Removes data points outside a specified range. Limits default to common sensor measurement limits.

5.  **Outlier Removal:** Remove outlier data points and mark questionable points. Removal can be done via a combination of automatic outlier detection and manual addition or removal of points.

6.  **Interpolation:** Fill in missing data points using interpolation. User can select between several different interpolation methods and can control how big of a gap is filled via interpolation.

7.  **Corrections:** Perform a shift correction for a selected set of points, apply a linear drift correction to a specific file, or apply a smoothing correction to a region. Drift corrections are based on calibration checks when provided.

8.  **fDOM Corrections:** Apply temperature and turbidity corrections to fDOM data using one of several methods. Methods default to published coefficient values but can be set to custom values.

9.  **Download Data:** Generate and view time summarized data; export data and metadata; and save the sonde project.

## Loading the App

To use the app run:

```{r load app, eval=FALSE, include=TRUE}
library(SondePolishR)
run_app()
```

------------------------------------------------------------------------

# Module Directions

## 1 Load Data

This module is used to load data into the app and is separated into five sections:

### 1.1 Sonde Project

This step is used to locate the datasets you want to use.

Use the **Browse** button to select:

-   An existing sonde project you have previously saved (`.RDS`).

-   One or more raw sonde data files (`.csv`). Note that currently only raw files from YSI EXO sondes are supported. For support of a different raw data format please submit an [issue request](https://github.com/wildfire-water-security/WWS-Node1-SondePolishR-sonde-qaqc/issues).

You can also load both an existing sonde project as well as new data you want to add to the project. In this case, any new data will be appended to the dataset.

### 1.2 Metadata

This step is used to locate the metadata you want to use.

Use the **Browse** button to select:

-   A field form file (`.csv`). This file is used to determine when the sonde was out of the water for maintenance or data downloads and view field notes during corrections. For details on the structure of the file see `help(example_fieldform)`.

-   A calibration check file (`.csv`). This file is used to visualize differences in the calibrations for the resident sonde and a freshly calibrated sonde. For details on the structure of the file see `help(example_calcheck)`.

### 1.3 Specify Site Info

This step is used to provide additional metadata about the sonde site.

-   **Enter a name or code for the site.** This will be used in the default file names when exporting the data.

-   **Specify the site time zone.** This will be used as the timezone for any DateTimes across the data and metadata.

Note that these are stored within the `sondeproj` so they only need to be input once per project.

### 1.4 Load Data

-   Before loading the data, supply an analyst name, this could be your initials or name. It will be used in the changelog to track who made what change.

-   This step uses the `load_project` function to read in data and format as a `sondeproj`. Use once you're specified the file paths to all applicable data and metadata.

-   To load a new project, use the **clear uploads** button to remove all previous file paths.

### 1.5 Add Precipitation

This step is used to add precipitation data to the project. Precipitation data is particularly helpful when correcting turbidity and fDOM data to determine if peaks are real or not. To add precipitation data there are three options:

1.  Provide a site latitude and longitude and download hourly precipitation data from [Merra-2](https://gmao.gsfc.nasa.gov/gmao-products/merra-2/) (uses the `get_precip` function).This dataset is available across a global scale at a resolution of 0.5 x 0.625 degrees available from 1981 to near real time.

2.  Provide a site latitude and longitude and download hourly precipitation data from [NLDAS](https://ldas.gsfc.nasa.gov/nldas) (uses the `get_precip` function). This dataset is available across CONUS at a resolution of 0.125 × 0.125 degrees available from 1981 to near real time. This data requires a token to access the data. See [here](https://urs.earthdata.nasa.gov/documentation/for_users/user_token) for directions on creating a token. Note that this token should be kept secret.

3.  Provide your own precipitation data as a `.csv` file. See `help(example_precip)` for the structure user input precipitation needs to have.

Once you've either input the site coordinates or specified the path to a precipitation file. Click **load precipitation** to add the data to the `sondeproj`.

------------------------------------------------------------------------

## 2 Visualize

This step is used to explore your dataset and offers a number of options to explore the data itself and it's associated metadata.

### 2.1 Explore Plot

**X-Axis**

You can control the x-axis (time) in the **Set Date Range** section.

-   Use the **date range** selector to specify a specific date range to view data

-   Toggle the **View data by period** button to click through periods of the data. Data periods will be based on the values of **date range**. Once you've turned on viewing by period, the buttons below the plot can be used to move between periods.

-   Adjust the **period length** to view different period lengths.

**Y-Axis**

In the top left of the page, you can select which variable you want to view as the primary variable. Changing this will change the variable shown in the plot on the right. You can also choose to show a secondary axis. Options for a secondary axis include:

-   **Raw Data:** The primary variable before any data corrections were made.

-   **Precipitation:** The hourly (or user supplied interval) precipitation. Only available if it was added to the project in **Load Data**.

-   **Other sonde variables:** You can also plot any of the other variables within the dataset.

Next to the plot you can also control the minimum and maximum y-values to display. It will default to the maximum value of the full dataset. You can also remove the value to have it autoscale to the data actively being viewed.

**Plot Options**

In addition to adjusting the axes, there are also options to include addition items on the plot or adjust coloring of the points:

-   **Plot points:** Displays data points.

-   **Plot lines:** Displays data as a line.

-   **Color points by file:** Color the data points by their file name.

-   **Show out-of-water periods:** Displays the out-out-water periods as shaded rectangles if metadata is provided.

-   **Show calibration checks:** Displays calibration check measurements if metadata is provided.

-   **Show questionable points:** Colors points marked as questionable in orange.

### 2.2 Explore Metadata

**Change Log**

This is the default table shown. Which is used to visualize the changes made to the data. As changes are made to the data, this table will get longer. The table can be filtered by values by clicking the while "All" box in each column. For example, you could view only the changes to fDOM. This is particularly useful as you can view changes to the data in the plot by clicking a row within the change log.

**Field Form**

This table shows the date of field visits with went the sonde was removed from the water. It also shows any field notes associated with the visit which can be useful when making decisions about corrections.

**Calibration Check**

This table shows the calibration checks performed for the parameter being plotted. Used serial numbers when the data was loaded, it determines if the probe was switched during the visit and estimates when the switch was made which is useful for drift corrections.

**Data Summary**

This table provides a summary of the data. The table is dynamic, and will summarize only the data being displayed in the above plot (note that it will not adjust when you use plotly's zoom tool, only the date range tools). This tool can be useful to get quick summary statistics about a period of interest like a storm event.

### 2.3 Correct Out of Water Periods

Out of water periods are periods where the sonde is still collecting measurements, but is not actively in the water due to maintenance or data downloads. These are clear periods where the data is not accurate and should be removed. This can be easily done using the **Flag OOW Periods** button which uses the `get_oow` function to identify periods where the sonde was out of the water and remove those points. Note that this function uses a 15-minute buffer on either side of the period to account for wiper test, inaccurate time reporting, and equilibration.

------------------------------------------------------------------------

## 3 Data Checks

This step is used to check, document, and resolve gaps and duplicated data within the dataset.

### 3.1 Gaps

Gaps are sections where no measurements were collected. These periods are identified using the `identify_gaps` function. Data gaps often occur when:

-   sonde batteries die during deployment

-   a sonde malfunctions

-   a sonde is temporarily removed for maintenance

If there are gaps within the dataset they will appear as a table showing the start, end, and length of the gap. The table is editable, so you can make a note of why there was a gap by double clicking the correct row in **User Notes**. This table is stored within the `sondeproj` and can be exported in the **Download Data** module.

### 3.2 Duplicates

Duplicates are periods where there is more than one observation at a given date and time. These periods are identified using the `identify_dups` function. Duplicates often occur when:

-   a sonde is switched out and deployment isn't immediately stopped

-   a sonde malfunctions

-   data is downloaded more than one time

Similar to gaps, this table is editable and will be saved within the project for export. However, duplicates can be particularly problematic particularly when interpolating. Additionally in the case of sonde switching, one set of data is known to be incorrect.

These duplicates can be resolved by clicking the row of the duplicate you want to view. It will pull up a plot of the duplicated region showing the duplicated values with the surrounding data. There are then options for resolving the duplicates:

-   **Use Mean:** This will average the values and replace the first duplicate with the averaged values.

-   **Remove Both:** When the correct dataset can't be determined, it may be best to remove the duplicated region and remove both sets.

-   **Use a specific set or file:** Sometime duplicated values are within the same file, in this case you can select which set to keep. Other times duplicated values are across different files, in this case you can select which file's values you want to keep. The values not selected are removed.

When resolving duplicates you can add a note which is stored within the change log (viewable in [2.2 Explore Metadata]) about why the duplicate was resolved the way it was. *Once the duplicate is resolved the table will refresh with the resolved duplicate removed.*

------------------------------------------------------------------------

## 4 Physical Limits

This step is used to remove points that are outside a set minimum and maximum. As it's name suggests, this is often used to remove points outside physical or measurement limits (i.e., a fDOM value below 0). The default values are based on reported measurement limits for YSI EXO probes. These can be manually adjusted in the **Set Physical Limits** section based on expert knowledge about your system. The limits are shown a red dashed lines within the plot and points falling outside the limits are colored red.

To save the limits and remove highlighted points outside the specified range, use the **Flag Points** section to add an optional note that will be saved in the change log and save the changes using the **Flag Points** button. The note in the change log by default will include the lower and upper limit ranges.

------------------------------------------------------------------------

## 5 Outlier Removal

This step is used to remove "bad" or erroneous data points. You can also mark regions or points that are questionable or bad. Points marked as bad are removed from the dataset while points marked as questionable are flagged but retained. There are several methods that can be used to select points for removal:

-   **None or manual selection only**: This will start with no points highlighted, use the selection tools with plotly to select and remove points to remove.

-   **Hampel filter:** Uses a [hampel filter](https://search.r-project.org/CRAN/refmans/pracma/html/hampel.html) to select outliers which is based on median absolute deviation across points. The selected points can be adjusted using the **k** parameter with controls how many points to include in the median calculation and **t** parameter which controls the threshold for selecting.

-   **Relative change:** Uses a relative percent difference across a median of **k** points. The **t** parameter controls the threshold for selecting.

-   **High variability:** Locates periods that have high variability relative to the surrounding data which may indicate unrealistic noise that should be removed.

All automatic selection methods can be used in combination with manual selection to adjust the selected points.

To remove the highlighted points, use the **Flag Points** section to add an optional note that will be saved in the change log and save the changes using the **Flag Points** button. The note in the change log by default will include the method used to select outliers. This step can be performed iteratively using different methods or including different user notes.

------------------------------------------------------------------------

## 6 Interpolation

This step is used to fill in missing data points either from data gaps or from data removals during corrections. You can set the maximum gap size that will be interpolated (max fill window) which prevent interpolating gaps that are too big. There are several methods available for interpolation which use the function `run_interp`:

-   **Linear:** Fills gaps using linear interpolation using the [`na.approx`](https://www.rdocumentation.org/packages/zoo/versions/1.8-15/topics/na.approx) function.

-   **Linear (seasonally adjusted):** Fills gaps using [`na.interp`](https://www.rdocumentation.org/packages/forecast/versions/9.0.1/topics/na.interp) function. This fills in gap using linear interpolation after adjusting for a seasonal temporal signal. The seasonal period can be adjusted using **season period** to adjust for different kinds of trends (i.e., daily fluxes, annual changes)

-   **Spline:** Fills gaps using cubic spline interpolation using the [`na.spline`](https://www.rdocumentation.org/packages/zoo/versions/1.8-15/topics/na.approx) function.

-   **Random Forest:** Uses the [`missForest`](https://www.rdocumentation.org/packages/missForest/versions/1.5/topics/missForest) function to fill gaps using a random forest model. The model Uses all available parameter values to fit a predictive model. Because of this, this method will likely not work well for regions where all parameters are missing.

To add the highlighted points, use the **Flag Points** section to add an optional note that will be saved in the change log and save the changes using the **Flag Points** button. The note in the change log by default will include the method used to interpolate with the fill window used. This step can be performed iteratively using different methods or including different user notes.

------------------------------------------------------------------------

## 7 Corrections

This step is used to apply shifts to sections of data. There are three types of corrections that can be applied. To switch between the methods, change the selection under **Select Type**.

### 7.1 Additive

This method is meant to correct groups of points that are shifted due to instrument issues (i.e., sensor interference). It works similarly to [5 Outlier Removal] where you use the plotly selection tools to select a group of points within the plot. Once you select a group of points, it will guess the slope and intercept needed apply a linear shift correction to fit the start and end points of the data. You can manually adjust these values using the **slope** and **intercept** inputs. For a absolute shift, you can set slope to 0 and adjust the intercept. **To clear the selected points, simply select a region of the plot with no points.**

### 7.2 Drift

This method is used to correct for drift in sensors over time. While not required, it works best when a calibration check file has been uploaded. The method works by file. When you select a file it will either use the calibration check data to determine the uncorrected (drifted) value and the corrected (freshly calibrated) value or guess based on the gap between the file and the next. The difference in values is used to apply a linear drift correction to the data and the corrected data is shown as a red line on the plot.

### 7.3 Smoothing

This method is used to correct for highly "messy" regions of data. The method works by applying a smoothing correction to a selected data range using the selection tool. Smoothing method include:

-   **rollmean**: Based on the `rollmean` function from the `zoo` package. Replaces data with a rolling mean value. The `k` parameter is used to control the number of points to include in the mean calculation.

-   **rollmedian**: Based on the `rollmedian` function from the `zoo` package. Replaces data with a rolling median value. The `k` parameter is used to control the number of points to include in the median calculation, should be odd but will convert to an odd number with a warning.

-   **kalman**: Based on the `dlmSmooth` function from the `dlm` package. Applies a Kalman filter which process model to the data. The `k` parameter is used controls the amount of smoothing.

This method should be used sparingly as it will "rewrite" or adjust large portions of data, including points you may have more confidence in.

For all methods, and similar to the other modules, when you want to save a change use the **Flag points** section. To add an optional note and save the updated values.

------------------------------------------------------------------------

## 8 fDOM Corrections

This step is used to correct fDOM data. fDOM is measured using an optical sensor that is affected by quenching due to temperature differences and turbidity interference. To determine the "true" fDOM value without these interferences, the data must be corrected. The corrected fDOM signal is shown as the red line on top of the data.

### 9.1 Temperature Corrections

[Watras et al. 2011](doi.org/10.4319/lom.2011.9.296) proposed a temperature correction equation of:

$$ fDOM_{T} = \frac{fDOM}{1 + \rho (T - 25)} $$

where $fDOM_T$ is the temperature corrected fDOM, $\rho$ is the correction factor and $T$ is the water temperature. The correction factor is known to vary somewhat by site. The default is 0.011 suggested by [Akie et al. 2024](doi.org/10.1002/hyp.70023).

### 9.2 Turbidity Corrections

Turbidity corrections can be quite complex and there have been a number of attempts at devising equations to correct for turbidity. The module currently supports five different correction options:

-   **None:** Skip turbidity corrections, more than temperature, turbidity corrections are thought to be site specific, particularly at high turbidities.

-   **Inverse Polynomial:** A equation suggested by [Fleck et al. 2026](https://doi.org/10.3133/ofr20261063) which applies a correction factor based on a quadratic equation with turbidity. This was found to fit similarly to more complex, five parameter models. Default values are the values reported by Fleck et al. 2026.

-   **Exponential (1-parameter):** One of the initial correction equations proposed by [Downing et al. 2012](doi.org/10.4319/lom.2012.10.767) which uses a single correction factor. The default correction factor comes from [Akie et al. 2024](doi.org/10.1002/hyp.70023) as the value for generic Elliot silt loam which was suggested to be appropriate to use for turbidity values *below* 300 FNU.

-   **Exponential (2-parameter):** A slightly modified version of the original 1-parameter model proposed by [Fleck et al. 2026](https://doi.org/10.3133/ofr20261063). This model works well for lower turbidity values but can have a poor fit at higher turbidity values (\>100-200 FNU). Default correction values come from [Fleck et al. 2026](https://doi.org/10.3133/ofr20261063).

-   **Exponential (5-parameter):** As an approach that would work across the full turbidity range, [Fleck et al. 2026](https://doi.org/10.3133/ofr20261063) proposed a 5-parameter model which fit high turbidity data well, but was complex to fit. Default correction values come from [Fleck et al. 2026](https://doi.org/10.3133/ofr20261063).

Similar to the other modules, when you want to save the fDOM correction use the **Flag points** section. To add an optional note and save the updated values.

------------------------------------------------------------------------

## 9 Download Data

This step is used to export cleaned data including data summaries and metadata and save the `sondeproj`. Similar to [1 Load Data], this step is separated into several sections.

### 9.1 Export Data

This section is used to view and export summaries of the data as `.csv` files. You have several options to control the data that is exported:

-   **Date Range:** Use the date selectors to only export data within the specified range.

-   **Export Frequency:** Aggregate the values into larger windows (e.g., hourly or daily values)

-   **Summary Method:** Available if **export frequency** is larger than the data interval. Used to specify the method used to summarize the data. You may select more than one and they will show up as separate column in the exported data.

> While there is an option to select a parameter, this only controls the data being shown in the plot. All of the parameters are exported to the file.

Once you've selected your data options, click **Choose Location** to set the file name and save location. Click **Export Data** to save the data to the file.

### 9.2 Export Metadata

This section is used to export metadata stored within the project to a `.csv` file. There are four different files that can be saved:

-   **Duplicate Notes:** Exports the table describing the data duplicates with any user added notes (see [3.2 Duplicates]).

-   **Missing Data Notes:** Exports the table describing the data gaps with any user added notes (see [3.1 Gaps]).

-   **Change Log:** Exports the change log describing all the changes made to the data.

-   **Precipitation:** Exports the hourly precipitation data downloaded from NASA power (see [1.5 Add Precipitation]).

### 9.3 Save Sonde Project

This section is used to save the `sondeproj` to a `.RDS` file (an R readable file). Similar to the other steps, use the **choose location** button to select the file location and name, then click **Export Project** to save the project to the file. This file can be reloaded into the app via [1.1 Sonde Project].

> Especially when doing more time intensive corrections it's highly recommended to save the project frequently to prevent loss of work if the app were to crash or bug out.
